Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold Standard Items

### 2025-08-01 manuscript.qmd
implement feedback

---
title: "Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold-Standard Items"
shorttitle: "Measuring latent variables with gold-standard items"
date: 2025-07-18
date-modified: 2025-07-30
date-format: iso
author:
  - name: Kathryn Eichhorn
    orcid: 0000-0003-3676-9420
    role: 
      - conceptualization
      - investigation
      - resources
      - data curation
      - writing
    affiliations:
      - name: University of the Bundeswehr Munich
        department: Institute of Psychology
        city: Neubiberg
        country: Germany
  - name: Markus Bühner
    corresponding: true
    email: buehner@lmu.de
    orcid: 0000-0002-0597-8708
    role: 
      - formal analysis
      - supervision
      - writing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
        address: Leopoldstr. 13
        city: D-80802 Munich
        country: Germany
        postal-code: 80802
  - name: Florian Pargent
    orcid: 0000-0002-2388-553X
    role:
      - formal analysis
      - software
      - validation
      - editing
    affiliations:
      - ref: lmu
  - name: Janika Saretzki
    orcid: 0000-0002-6536-8266
    role:
      - editing
    affiliations:
      - name: University of Graz
        department: Department of Psychology
        city: Graz
        country: Austria
      - name: Charlotte Fresenius University of Psychology
        city: Munich
        country: Germany
      - ref: lmu
  - name: Larissa Sust
    orcid: 0000-0002-3389-1626
    role:
      - editing
    affiliations:
      - ref: lmu
  - name: Jonas Hauck
    orcid: 0009-0002-8872-0530
    role:
      - editing
    affiliations:
      - id: "regen"
        name: University of Regensburg
        department: Faculty of Psychology
        city: Regensburg
        country: Germany
  - name: Sven Hilbert
    orcid: 0000-0001-5808-8357
    role:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: regen

author-note:
  disclosures:
    data-sharing: This is version 0 (last modification {{< meta date-modified >}}) of our preprint to be published on PsyArXiv. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Kathryn Eichhorn and Markus Bühner contributed equally to this manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on cognitive interviews ($N$ = `r nrow(dat_interviews)`) and a pilot study ($N$ = `r nrow(dat_pilot)`) with an initial item set, three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two ($N$ = `r nrow(dat_dich)`) or six ($N$ = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

One important aspect of item construction is item wording. In this context, @pargentCanMakeIt2019 showed that the quality of item wording did, in fact, not impact model fit based on commonly used fit indices of confirmatory factor analyses. Sound item wording may not be crucial for achieving good model fit, but it clearly is for understanding the item correctly: For example, if a person does not correctly recognize an item with negative polarity, the person's value on the scale is immediately over- or underestimated by a few points. In some cases, this can lead to considerable distortions in the test score. If a person does not understand the items correctly, their response cannot be valid, regardless of any model fit indicators [for an extreme example, see @maulRethinking2017]. Thus, it would be problematic to evaluate the quality of a questionnaire by solely looking at the model fit of the latent variable model, the discrimination parameters and/or reliability estimates. In this context, @mcneishThornyRelationMeasurement2018 stated that good fit indices (e.g., Root Mean Square Error of Approximation, RMSEA; Standardized Root Mean Square Residual, SRMR; and the Comparative Fit Index, CFI) must be considered together with measurement quality (reliability) and the size of standardized factor loadings. They showed examples where all fit indices were excellent, but the standardized item loadings were .40, and another model where the fit indices were inadequate, but the factor loadings were .90. Thus, fit indices can be misleading without considering the quality of measurement.

@@

The total sample of our main study consisted of `r nrow(dat_desc)` participants aged between `r min(dat_desc$Alter_realW)` and `r max(dat_desc$Alter_realW)` years (*M* = `r printnum(mean(dat_desc$Alter_realW))`, *SD* = `r printnum(sd(dat_desc$Alter_realW))`), including `r nrow(dat_desc[dat_desc$Geschlecht == 1,])` women (`r round(nrow(dat_desc[dat_desc$Geschlecht == 1,]) / nrow(dat_desc) * 100)`%) and `r nrow(dat_desc[dat_desc$Geschlecht == 2,])` men (`r round(nrow(dat_desc[dat_desc$Geschlecht == 2,]) / nrow(dat_desc) * 100)`%). The overall sample consisted of two independent subsamples, which were divided into the 2-point and 6-point response format conditions.

@@

@tbl-1 shows the distribution of gender, level of education, native language, and age within the two conditions. As expected based on the randomization procedure, the two subsamples are very similar. Due to the left-skewed age distribution and outlier values, the median is reported as a measure of the central tendency instead of the mean.

@@

As expected, reliability estimates were higher for scales with six response categories compared to two categories. The results with regard to model fit and validity, indicated as correlations between latent variables compared to the correlation between gold-standard items, are mixed. Some fit indices suggested slightly better fit for items with two response categories and some for six categories. The same holds true for the differences between the correlations of the gold-standard items and the correlations of the latent variables. Except for reliability, the study revealed no clear picture to choose between two or six response categories. Hilbert et al. [-@hilbertInfluenceResponseFormat2016; -@hilbertWhatMeasureEmpirical2022] showed that the same items with a different number of response categories do not measure the same latent variable. This result highlights that the choice of the number of response categories constructing a questionnaire should be based on an empirical study comparing several options accompanied by an intensive cognitive pretesting phase and not only based on the literature which focuses on maximal reliability.

>>>>

---
title: "Item Construction Rules Revisited: Learnings from Measurement of Latent Variables with Gold-Standard Items"
shorttitle: "Measuring latent variables with gold-standard items"
date: 2025-07-18
date-modified: 2025-07-30
date-format: iso
author:
  - name: Kathryn Eichhorn
    orcid: 0000-0003-3676-9420
    role: 
      - conceptualization
      - investigation
      - resources
      - data curation
      - writing
    affiliations:
      - name: University of the Bundeswehr Munich
        department: Institute of Psychology
        city: Neubiberg
        country: Germany
  - name: Markus Bühner
    corresponding: true
    email: buehner@lmu.de
    orcid: 0000-0002-0597-8708
    role: 
      - formal analysis
      - supervision
      - writing
    affiliations:
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
        address: Leopoldstr. 13
        city: D-80802 Munich
        country: Germany
        postal-code: 80802
  - name: Florian Pargent
    orcid: 0000-0002-2388-553X
    role:
      - formal analysis
      - software
      - validation
      - editing
    affiliations:
      - ref: lmu
  - name: Janika Saretzki
    orcid: 0000-0002-6536-8266
    role:
      - editing
    affiliations:
      - name: University of Graz
        department: Department of Psychology
        city: Graz
        country: Austria
      - name: Charlotte Fresenius University of Psychology
        city: Munich
        country: Germany
      - ref: lmu
  - name: Larissa Sust
    orcid: 0000-0002-3389-1626
    role:
      - editing
    affiliations:
      - ref: lmu
  - name: Jonas Hauck
    orcid: 0009-0002-8872-0530
    role:
      - editing
    affiliations:
      - id: "regen"
        name: University of Regensburg
        department: Faculty of Human Sciences
        city: Regensburg
        country: Germany
  - name: Sven Hilbert
    orcid: 0000-0001-5808-8357
    role:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: regen

author-note:
  disclosures:
    data-sharing: This is version 0 (last modification {{< meta date-modified >}}) of our preprint to be published on PsyArXiv. All materials (reproducible manuscript, analysis code, datasets, codebooks, questionnaires) are available in the project's repository on the Open Science Framework (OSF) at <https://osf.io/p7492/>. A Quarto Manuscripts website is hosted at <https://florianpargent.github.io/gold-standard-items/>. The data has previously been analyzed in the dissertation thesis of the first author at <https://doi.org/10.5282/edoc.23628>. As part of the dissertation, a preregistration was published at <https://osf.io/cz3uv/?view_only=9cac02db231b48629fea9ae53c3038b9> that differs from the analyses reported in the current manuscript.
    conflict-of-interest: The authors declare that there were no conflicts of interest concerning the authorship or the publication of this article.
    financial-support: ~
    gratitude: ~
    authorship-agreements: ~
language:
  title-block-role-introduction: "Kathryn Eichhorn and Markus Bühner contributed equally to this manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; <https://credit.niso.org/>):"
abstract: "This article investigates whether gold-standard items are helpful for questionnaire construction. Based on an initial item pool (138 items) derived by a deductive item construction ($N$ = 124) and prototype approach ($N$ = 24), followed by cognitive interviews ($N$ = `r nrow(dat_interviews)`) and a pilot study ($N$ = `r nrow(dat_pilot)`) of a preliminary item set (61 items), three 12-item scales were constructed to measure the physical traits of body height, body weight, and age. We collected data on these scales using response formats with either two ($N$ = `r nrow(dat_dich)`) or six ($N$ = `r nrow(dat_poly)`) categories. We also collected numeric self-reports of body weight, body height, and age as gold-standard items. Confirmatory factor analyses revealed that the gold-standard items did not consistently exhibit the highest loading on their corresponding latent variable. Furthermore, when controlling for the self-reported physical body height, body weight, and age as gold-standard items, as well as gender, we did not always find an interpretable, systematic residual variance. Finally, the pattern of correlations between the latent variables did not reflect the correlations between the self-reported gold-standard items, suggesting that the item scales and the gold-standard items do not have the same validity. While these results are consistent with previous studies, our analyses also showed that items with two response categories were at least as valid as those with six categories, contradicting past recommendations. When constructing a questionnaire, we would argue that the items intended to measure the latent variable most directly should have the highest loading on that variable. If this is not the case, the content validity is questionable at best. The implications are that intensive cognitive pretesting is necessary. Questionnaires with different response formats should be compared empirically and hypotheses about which items best represent the latent variable should be tested."
license: CC BY 4.0
bibliography: [references.bib, r-references.bib]
floatsintext: true
numbered-lines: true
csl: apa.csl
---

@@

One important aspect of item construction is item wording. In this context, @pargentCanMakeIt2019 and @goretzkoNotVeryPowerful2020 showed that the quality of item wording did, in fact, not impact model fit based on commonly used fit indices of confirmatory factor analyses. Sound item wording may not be crucial for achieving good model fit, but it clearly is for understanding the item correctly: For example, if a person does not correctly recognize an item with negative polarity, the person's value on the scale is immediately over- or underestimated by a few points. In some cases, this can lead to considerable distortions in the test score. If a person does not understand the items correctly, their response cannot be valid, regardless of any model fit indicators [for an extreme example, see @maulRethinking2017]. Thus, it would be problematic to evaluate the quality of a questionnaire by solely looking at the model fit of the latent variable model, the discrimination parameters and/or reliability estimates. In this context, @mcneishThornyRelationMeasurement2018 stated that good fit indices (e.g., Root Mean Square Error of Approximation, RMSEA; Standardized Root Mean Square Residual, SRMR; and the Comparative Fit Index, CFI) must be considered together with measurement quality (reliability) and the size of standardized factor loadings. They showed examples where all fit indices were excellent, but the standardized item loadings were .40, and another model where the fit indices were inadequate, but the factor loadings were .90. Thus, fit indices can be misleading without considering the quality of measurement.

@@

The total sample of our main study consisted of `r nrow(dat_desc)` participants aged between `r min(dat_desc$Alter_realW)` and `r max(dat_desc$Alter_realW)` years (*M* = `r printnum(mean(dat_desc$Alter_realW))`, *SD* = `r printnum(sd(dat_desc$Alter_realW))`), including `r nrow(dat_desc[dat_desc$Geschlecht == 1,])` women (`r round(nrow(dat_desc[dat_desc$Geschlecht == 1,]) / nrow(dat_desc) * 100)`%) and `r nrow(dat_desc[dat_desc$Geschlecht == 2,])` men (`r round(nrow(dat_desc[dat_desc$Geschlecht == 2,]) / nrow(dat_desc) * 100)`%). The overall sample consisted of two independent subsamples, which were divided into the 2-point ($N$ = `r nrow(dat_desc[dat_desc$categories == "2 categories",])`) and 6-point ($N$ = `r nrow(dat_desc[dat_desc$categories == "6 categories",])`) response format conditions.

@@

@tbl-1 shows the distribution of gender, level of education, native language, and age within the two conditions. As expected based on the randomization procedure, the two subsamples are very similar. Due to the left-skewed age distribution and outlier values, we report both the mean and the median.

@@

As expected, reliability estimates were higher for scales with six response categories compared to two categories [@prestonOptimalNumberResponse2000]. The results with regard to model fit and validity, indicated as correlations between latent variables compared to the correlation between gold-standard items, are mixed. Some fit indices suggested slightly better fit for items with two response categories and some for six categories. The same holds true for the differences between the correlations of the gold-standard items and the correlations of the latent variables. Except for reliability, the study revealed no clear picture to choose between two or six response categories. Hilbert et al. [-@hilbertInfluenceResponseFormat2016; -@hilbertWhatMeasureEmpirical2022] showed that the same items with a different number of response categories do not measure the same latent variable. This result highlights that the choice of the number of response categories constructing a questionnaire should be based on an empirical study comparing several options accompanied by an intensive cognitive pretesting phase and not only based on the literature which focuses on maximal reliability.
